FAIR fission track analysis with geochron@home

### 2025-07-26 geochron@home/manuscript/paper.tex
Added summary tables to geochron@home paper

geochron@home serves four different purposes. It can be used (1) to
count fission tracks in `private mode', i.e. hidden from other users
on the internet; (2) to archive fission track images and counts for
inspection by other users; (3) to create tutorials for new students of
the fission track method; and (4) to generate crowd-sourced fission
track data, by serving randomly selected selections of images to
citizen scientists. We illustrate these four applications with
examples that demonstrate (1) geochron@home's ability to compare and
combine fission track counts for multiple users within a lab group;
(2) the value of the geochron@home archive in the peer review system;
(3) the use of simple tutorials in teaching novice users how to count
fission tracks; and (4) the `wisdom of crowds' in fission track
identification.\medskip

@@

We anticipate that the introduction of FAIR workflows will make
fission track data more accurate and more future proof. Storing
fission track data online will benefit future developments in fission
track thermochronology. For example, archival datasets of peer
reviewed fission track counts can be used to train and improve machine
learning algorithms for automated fission track analysis. We invite
other geochronological methods to follow the fission track community's
lead in FAIR data processing. This would benefit all the other Earth
Science disciplines that depend on geochronological data.
\end{abstract}

@@

The combined data plots of Figures~\ref{fig:AvP}b and c contain two
sets of counts, with $\sum N_{1} = 686$ and $\sum N_{2} = 679$, and a
weighted mean track density ratio $\hat{\rho}_{1}/\hat{\rho}_{1} =
0.94$ with relative standard error 0.013. PV counted 600 tracks in
those common areas $A_0$, of which 549 were also counted by
AC. Conversely, AC counted 622 tracks of which 549 were also counted
by PV. The ratio of the two analysts' track density estimates based
just on the common area is therefore 600/622 = 0.965 with relative
standard error 0.016, calculated from Equation~\ref{eq:NO}. This is
close to the weighted mean ratio of 0.94 (Figure~\ref{fig:AvP}c), and
is slightly nearer to 1. It indicates that PV under-counts the Mount
Dromedary apatite by 3.5\% relative to AC.  The existence of `observer
bias' is well documented \citep{tamer2025}.  It is one of the reasons
why fission track analysis is often done relative to age standards:
observer bias does not have to be a problem provided that it is
consistent between grains, and between samples.\medskip

@@

Figure~\ref{fig:radialcrowd}a shows box plots and summary statistics
for the raw counts made by 68 students (two of the students were
unable to complete the assignment) for the same selection of 25 grains
that were analysed by AC and PV in Section~\ref{sec:private}. Everyone
counted tracks in the same ROIs, and counts made by PV for those ROIs
have been added as blue dots.  Variation between the mean values is to
be expected because of the differing numbers of tracks due to varying
areas and U contents of the grains. But the variation between counts
within grains (shown by the standard deviations) is due entirely to
differences between the students' recognising and counting exactly the
same tracks. These standard deviations increase with the mean number
of tracks and are considerable in size, being on average about 40\% of
the mean.  Also, the vast majority of students counted fewer tracks
than PV did, often many fewer. In 21 of the 25 grains, PV's count is
above the upper quartile of the students', and on two occasions
someone counted no tracks at all.  With expert trained counters one
would expect much smaller differences between counts.  Nevertheless
there are many repeated counts also.\medskip

Figure~\ref{fig:radialcrowd}b shows a scatter plot of pairs of counts
for two grains made by 37 students who counted both of them. They
average around 29 and 30 tracks per grain compared with PV's pair of
52 and 50, which is highlighted in blue. There is a positive
correlation between the pairs, consistent with systematic observer
effects (i.e., people who count a low value in one grain tend to count
similarly low in the other, and vice versa).  Comparing other pairs of
grains shows similar results, with correlations varying between 0.3
and 0.9. In all cases we have looked at, the residual variation, after
allowing for systematic differences between students, is consistent
with Poisson variation.  Figure~\ref{fig:radialcrowd}c shows a radial
plot of the ratios of the counts in Panel~b, using the precisions
given by
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The context here is rather different from that in
Figure~\ref{fig:AvP}b, where we were comparing the ratios of counts
made by two analysts on each of 25 grains. Here we are comparing the
ratios of counts for the same two grains made by each of 37
analysts. There is no theoretical justification for
Equation~\ref{eq:N12} here, but it still provides a useful benchmark
for assessing variation between the ratios. In
Figure~\ref{fig:radialcrowd}c, that variation is consistent with
Poisson variation (as it is for most other pairs of grains) but with
expert counters one would expect less variation than that. However, in
spite of the under-counting by the students, the pooled students'
ratio of 0.76 is close to PV's ratio of 0.70. This happens quite
often in this data set, but not always, and there are some pairs of
grains where the students' pooled ratio differs somewhat from
PV's.\medskip

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:crowdsourcing} is a two-dimensional histogram for the
x-y positions of all the track positions generated by the students in
the two apatite crystals. Visual comparison of the histogram with the
optical image confirms that the students unanimously identified the
most obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets.\medskip

Figure~\ref{fig:crowdsourcing} also shows that some students counted
the tails of the fission tracks rather than their etch pits, despite
being told the opposite in the tutorial. Fixing this issue will
require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

@@

%\appendix
%\section{}

%\appendix{}

>>>>

geochron@home serves four different purposes. It can be used (1) to
count fission tracks in `private mode', i.e. hidden from other users
on the internet; (2) to archive fission track images and counts for
inspection by other users; (3) to create tutorials for new students of
the fission track method; and (4) to generate crowd-sourced fission
track data, by serving randomly selected selections of images to
citizen scientists. We illustrate these four applications with
examples that demonstrate (1) geochron@home's ability to compare and
combine fission track counts for multiple users within a lab group;
(2) the value of the geochron@home archive in the peer review system;
(3) the use of simple tutorials in teaching novice users how to count
fission tracks; and (4) the `wisdom of crowds' in fission track
ratio estimation.\medskip

@@

We anticipate that the introduction of FAIR workflows will make
fission track data more accurate and more future proof. Storing
fission track data online will benefit future developments in fission
track thermochronology. For example, archival datasets of peer
reviewed fission track counts can be used to train and improve machine
learning algorithms for automated fission track analysis. We invite
other geochronological methods to follow the fission track community's
lead in FAIR data processing. This would benefit all the Earth Science
disciplines that depend on geochronological data.
\end{abstract}

@@

The combined data plots of Figures~\ref{fig:AvP}b and c contain two
sets of counts, with $\sum N_{1} = 686$ and $\sum N_{2} = 679$, and a
weighted mean track density ratio $\hat{\rho}_{1}/\hat{\rho}_{1} =
0.94$ with relative standard error 0.013. PV counted 600 tracks in
those common areas $A_0$, of which 549 were also counted by
AC. Conversely, AC counted 622 tracks of which 549 were also counted
by PV. The ratio of the two analysts' track density estimates based
just on the common area is therefore 600/622 = 0.965 with relative
standard error 0.018, calculated from Equation~\ref{eq:NO}. This is
close to the weighted mean ratio of 0.94 (Figure~\ref{fig:AvP}c), and
is slightly nearer to 1. It indicates that PV under-counts the Mount
Dromedary apatite by 3.5\% relative to AC.  The existence of `observer
bias' is well documented \citep{tamer2025}.  It is one of the reasons
why fission track analysis is often done relative to age standards:
observer bias does not have to be a problem provided that it is
consistent between grains, and between samples.\medskip

@@

We here present the results of a type of `crowd-sourcing' experiment
that was carried out at UCL as part of an undergraduate course in
isotope geology. 70 students were asked to create an account on the
geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains.  This
large dataset was parsed into separate data files --- one for each
student --- which they processed during an assessed practical
exercise.\medskip

The raw counts for 68 students (two of the students were unable to
complete the exercise) are given in Table~\ref{tab:crowdtable} of the
Appendix, for the same selection of 25 grains that were analysed by AC
and PV in Section~\ref{sec:private}, along with corresponding counts
made by PV. Everyone counted tracks in the same ROIs.
Figure~\ref{fig:radialcrowd}a shows box plots and summary statistics
of these counts for each grain.  Variation between the mean values is
to be expected because of the differing numbers of tracks due to
varying areas and U contents of the grains. But the variation between
counts within grains (shown by the standard deviations) is due
entirely to differences between the students’ recognising and counting
exactly the same tracks. These standard deviations increase with the
mean number of tracks and are considerable in size, being on average
about 40\% of the mean. With expert trained counters one would expect
much smaller differences between counts.  Furthermore, the vast
majority of students counted fewer tracks than PV did, often many
fewer, and on two occasions someone counted no tracks at all. PV's
count is always above the students' median and nearly always above the
upper quartile.  Table~\ref{tab:crowdtable} (looking along rows) shows
just 7 students with average counts within 10\% of PV's, and 19 with
average counts less than half of his.\medskip

Figure~\ref{fig:radialcrowd}b shows a scatter plot of pairs of counts
for two grains made by 36 students who counted both of them. They
average around 29 and 30 tracks per grain compared with PV's pair of
52 and 50, which is highlighted in blue. There is a positive
correlation between the pairs, consistent with systematic observer
effects (i.e., people who count a low value in one grain tend to count
similarly low in the other, and vice versa).  Comparing other pairs of
grains shows similar results, with correlations varying between 0.4
and 0.9. In all cases we have looked at, the residual variation, after
allowing for systematic differences between students, is consistent
with Poisson variation.  Figure~\ref{fig:radialcrowd}c shows a radial
plot of the ratios of the counts in Panel~b, using the precisions
given by
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The context here is rather different from that in
Figure~\ref{fig:AvP}b, where we were comparing the ratios of counts
made by two analysts on each of 25 grains. Here we are comparing the
ratios of counts for the same two grains made by each of 37
analysts. There is no theoretical justification for
Equation~\ref{eq:N12} here, but it still provides a useful benchmark
for assessing variation between the ratios. In
Figure~\ref{fig:radialcrowd}c, that variation is consistent with
Poisson variation (as it is for most other pairs of grains) but with
expert counters one would expect less variation than that. However, in
spite of the under-counting by the students, the pooled students'
ratio of 0.97 is close to PV's ratio of 1.04. This happens quite often
in this data set, but not always, and there are some pairs of grains
where the students' pooled ratio differs somewhat from PV's.  The
logcontrast between the student ratios and PV's ratios for all 300
pairs of grains follows a normal distribution with a mean of 0.0 and
standard deviation 0.23 (Table~\ref{tab:logcontrasts}). This indicates
that, on the whole, the student estimates of the (log)ratios are
unbiassed with respect to the expert opinion of PV.
\medskip

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:crowdsourcing} is a two-dimensional histogram for the
x-y positions of all the track positions generated by the students in
the two apatite crystals. Visual comparison of the histogram with the
optical image confirms that the students unanimously identified the
most obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets. Figure~\ref{fig:crowdsourcing} also shows that some students
counted the tails of the fission tracks rather than their etch pits,
despite being told the opposite in the tutorial. Fixing this issue
will require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

@@

\appendix
\section{Summary tables for the crowd-sourcing experiment}

\begingroup
\captionof{table}{ Summary table of the crowd sourcing results for the
  different users (rows) and grains (columns).  The last row is the
  ratio of the median count to PV's count (user~1) for each grain.
  The last column is the ratio of the weighted mean count over grains
  compared with PV for the same grains.
}
\label{tab:crowdtable}
\renewcommand{\arraystretch}{0.7}
\begin{scriptsize}
  \input{../output/crowdtable.txt}
\end{scriptsize}
\endgroup

\clearpage

\begingroup
\captionof{table}{
  Logcontrasts between the pooled count ratio of the
  students and PV for each pair of 25 grains in
  Table~\ref{tab:crowdtable}.
}
\label{tab:logcontrasts}\medskip
\renewcommand{\arraystretch}{0.7}
\begin{scriptsize}
  \input{../output/logcontrasts.txt}
\end{scriptsize}
\endgroup
